High geomagnetic field intensity recorded by anorthosite xenoliths requires a vigorous late Mesoproterozoic geodynamo

### 2021-11-09 PNAS_manuscript/PNAS_manuscript.tex
manuscript update

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution history of Earth's core. However, such records are often difficult to recover due to the alterations associated with weathering after the formation of rocks. In this study, we use plagioclase cumulates from the North American Midcontinent Rift that has experienced minimal post-formation alterations to obtain a record of $\sim$89 ZAm$^2$ virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\authordeclaration{The authors declare no conflict of interest.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{2}To whom correspondence should be addressed. E-mail: yimingzhang\@berkeley.edu}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{absolute paleointensity $|$ Laurentia $|$ Midcontinent Rift $|$ anorthosite $|$ geodynamo $|$}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first principles calculations implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma) \cite{Labrosse2015a, Ohta2016a}. With the estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of the core and mantle.

@@

Paleomagnetic data from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Previous interpretations of the current Precambrian absolute paleointensity records (compiled in the PINT database; \url{http://earth.liv.ac.uk/pint/}) suggest that an appreciable geomagnetic field existed in the Archean and progressively decayed through the Proterozoic till it reached its weakest in the Ediacaran when very low virtual dipole moment (VDM) estimates have been obtained \cite{Bono2019a, Shcherbakova2019a, Thallner2021b, Lloyd2021a}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Bono et al \cite{Bono2019a} adopted a billion-year-scale view of such observations of the Precambrian geomagnetic field intensities to suggest that these evidence are consistent with models that use high core thermal conductivity values where the nucleation of the inner core began to provide compositional convection power for driving the geodynamo during the Ediacaran-Cambrian after experiencing a progressive decay when the geodynamo was dominantly driven by thermal conduction earlier in the Proterozoic \cite{Driscoll2016a, Davies2021a}.

@@

However, previous paleointensity data from intrusive and extrusive rocks in the \textit{ca.} 1.1 Ga Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VDM values similar to that of today, which are higher than the values predicted by the long term decay curve proposed by Bono et al \cite{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Pesonen and Halls \cite{Pesonen1983b} obtained an average virtual dipole moment of 93 $\times 10^21$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks from the Midcontinent Rift, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flows of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). Kulakov et al \cite{Kulakov2013a} obtained an average VDM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by ref. \citealp{Pesonen1983b, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Pesonen and Halls \cite{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas Kulakov et al \cite{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by Sprain et al \cite{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VDM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}. This high value from the Midcontinent Rift volcanics may be incompatible with the hypothesis of a ``Proterozoic dipole low" \cite{Biggin2009a}.

Although it is typical of paleointensity experiments to suffer from low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from ``IZZI" modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the anorthosite xenoliths have a very high success rate. Their consistent specimen- and site-level paleointensity results together with rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. With more single-slope, high quality paleointensity data, further enrich the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

The Mesoproterozoic North American Midcontinent Rift is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; \cite{Vervoort2007a, Miller2013a}) was characterized by rapid and voluminous emplacement of intrusive rocks and extrusive volcanics. The $<$500 Kyr emplacement of the \textit{ca.} 1096 Ma Duluth Complex and the $~$8 km thick volcanics associated with it is one manifestation \cite{Swanson-Hysell2021a}. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex punctuates another such period of magmatism during the Midcontinent Rift main stage activity. High precision $^{206}$Pb/$^{238}$U zircon dates from one anorthosite xenolith and units that have cross-cutting relationships with the Beaver River diabase tightly constrain the timing of diabase emplacement to be at 1091.7 $\pm$ 0.2 Ma \cite{Zhang2021b}. The xenolith-bearing diabase dikes and sills intruded into hypabyssal depths via wide conduits that accommodated the entrainment of numerous anorthosite xenoliths that have short-axis diameters ranging from centimeters to up to 180 meters \cite{Boerboom2004a, Boerboom2006b}.

Previous geochronology data and rare earth element diffusion modeling suggest that the anorthosite xenoliths of the Beaver River diabase are cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Petrological textures and geochemical attributes imply that the Beaver River anorthosite xenoliths formed in a lower crustal magmatic condition distinct from that of the Duluth Complex which is $\sim$4 Myr older and also generated massive anorthositic cumulates that led to the formation of the Anorthositic Series \cite{Zhang2021b, Swanson-Hysell2021a}. The dominantly monomineralic anorthosite xenoliths of the Beaver River diabase have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. The plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture showing igneous foliations (Fig. \ref{Petro_QDM}A), but the more calcium-rich plagioclase of the Beaver River anorthosites often develop granoblastic texture characterized by equigranular crystals with poor alignments (Fig. \ref{Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial symplectic titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex Anorthositic Series rocks \cite{Miller1990a}. Such a high concentration of Fe oxides carrying magnetic remanence in the slowly cooled Duluth Complex anorthositic rocks reflect that iron in the plagioclase mush in the magma chamber acted as a means of buffering the melt with respect to oxygen, forming crystallites along preexisting plagioclase substrate or as iron oxides incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures \cite{Scofield1986a}. In contrast, titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent but relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}). This likely reflects that the Beaver River diabase magma had a different magma oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

% the Beaver River diabase rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers and likely did not reach  \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \cite{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}).

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from \cite{Pesonen1983a} are categorized by lithology. All sites from \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; \cite{Zhang2021b}). Petrographic image (A) shows Duluth Complex anorthosite have euhedral, interlocking plagioclase crystals showing igneous foliation. (B) Beaver River anorthosite have plagioclase with a granoblastic texture. (C) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) at UC Berkeley Paleomagnetism Lab show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the complex overlapping magnetic fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure*}

@@

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, yielding 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots 
\cite{Arai1963a} are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table S1 and are plotted in Fig. \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments displayed on Arai and Zijderveld plots (insets) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent our fitting for the dominant single-slope component that passes our acceptance criteria with estimates of the ancient field (B$_{anc}$) being used to determine site means. The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show typical non-ideal sagging and zigzagging behaviour that fails our acceptance criteria. Plot for diabase specimen BD2-5a shows typical paleointensity behavior of this site - specimens typically have double-slope behaviors although the lower temperature slopes can pass the selection criteria, resulting in the high estimated B$_{anc}$ values shown in Fig. \ref{fig:PINT_cooling_corrected}. Thus, the estimates of B$_{anc}$ for this specimen plot is illustrative, but were not used to calculate a site mean paleointensity value in Fig. \ref{fig:PINT_cooling_corrected}.}}
\label{fig:IZZI_examples}
\end{figure*}

>>>>

% Please add a significance statement to explain the relevance of your work
\significancestatement{Acquiring high-fidelity observational records of ancient magnetic field intensity from remanent magnetization of rocks is crucial for constraining the evolution history of Earth's core. However, such records are often difficult to recover due to the alterations associated with weathering after the formation of rocks and heating during experiments. In this study, we use plagioclase cumulates from the North American Midcontinent Rift that has experienced minimal post-formation alterations to obtain a present-day-like virtual dipole moment 1.1 billion years ago. The results indicate the existence of strong power sources for Earth's dynamo during the late Mesoproterozoic. }

@@

\authordeclaration{The authors declare no conflict of interest.}
% \equalauthors{\textsuperscript{1}A.O.(Author One) contributed equally to this work with A.T. (Author Two) (remove if not applicable).}
\correspondingauthor{\textsuperscript{1}To whom correspondence should be addressed. E-mail: yimingzhang\@berkeley.edu.}

% At least three keywords are required at submission. Please provide three to five keywords, separated by the pipe symbol.
\keywords{absolute paleointensity $|$ Laurentia $|$ Midcontinent Rift $|$ anorthosite $|$ geodynamo $|$ Mesoproterozoic $|$ inner core}

@@

% If your first paragraph (i.e. with the \dropcap) contains a list environment (quote, quotation, theorem, definition, enumerate, itemize...), the line after the list may have some extra indentation. If this is the case, add \parshape=0 to the end of the list environment.
\dropcap{E}arth's magnetic field is the result of convective flow of liquid iron-alloy in Earth's outer core. At present day, the geodynamo is collectively powered by heat flow across the core-mantle boundary (CMB), and from the crystallization of the solid inner core from the liquid outer core which provides latent heat and compositional buoyancy. However, while paleomagnetic studies have found that a dynamo field has existed since at least 3.4 billion years ago in the Archean \cite{Selkin2007a, Tarduno2014a, Brenner2020a}, Earth's inner core likely crystallized more recently. Estimates of the timing of the initial crystallization of Earth's inner core are interconnected with estimates for the core's thermal conductivity. Higher conductivity values imply faster cooling rates, which in turn imply that the threshold for the freezing of the inner core happened more recently \cite{Davies2015a}. While some estimates of these values are consistent with an inner core age $>$3 Ga \cite{Gubbins2004a, Konopkova2016a}, other estimates based on first principles calculations implied higher thermal conductivity values and an age for the inner core that is less than 1.3 Ga \cite{Pozzo2012a, Koker2012a, Gomi2013a, Zhang2020b}, with some suggesting even younger ages ($<$ 700 Ma; ref.s \citealp{Labrosse2015a, Ohta2016a}. With the estimates for the core's thermal conductivity continue to be debated, it is crucial to use observational records as an independent constraint on the thermal evolution of Earth's core and mantle.

@@

Paleomagnetic records from ancient rocks are one of the few types of observational data that have the potential to record the evolution of Earth’s core and can be used to test the predictions from numerical simulations. Interpretations of selected records of the current Precambrian absolute paleointensity database (compiled in the PINT database; \url{http://earth.liv.ac.uk/pint/}) suggest that an appreciable geomagnetic field existed in the Archean but reached its weakest in the Ediacaran when very low virtual dipole moment (VDM) estimates have been obtained \cite{Bono2019a, Shcherbakova2019a, Thallner2021b, Lloyd2021a}, followed by a recovery near the Ediacaran-Cambrian boundary \cite{Thallner2021a}. Ref. \citealp{Bono2019a} adopted a billion-year-scale view of such observations to suggest that these data are consistent with thermal evolution models that use high core thermal conductivity values, where the nucleation of the inner core did not begin to provide compositional convection power for driving the geodynamo until the Ediacaran-Cambrian, suggesting that the Proterozoic dynamo might have experienced a progressive decay when compositional convection was lacking and thermal convection was decreasing \cite{Driscoll2016a, Davies2021a}.

@@

However, previous paleointensity data from intrusive and extrusive rocks in the \textit{ca.} 1.1 Ga Midcontinent Rift \cite{Pesonen1983a, Kulakov2013a, Sprain2018a} record VDM values similar to that of today, which are higher than the values predicted by the long term decay curve proposed by ref. \citealp{Bono2019a} (Fig. \ref{fig:PINT_compilation}). Ref. \citealp{Pesonen1983a} obtained an average virtual dipole moment of 93 $\times 10^{21}$ Am$^2$ (ZAm$^2$) after applying a 13\% cooling rate correction (114 ZAm$^2$ before correction) using both extrusive and intrusive rocks from the Midcontinent Rift, including the \textit{ca.} 1108-1105 Ma Osler Volcanics, the \textit{ca.} 1100 Ma Maimainse Point volcanics, the \textit{ca.} 1107 Ma intrusive Logan Sills, Thunder Bay dikes, the Baraga and Marquette dikes, as well as lava flows of the \textit{ca.} 1096 Ma North Shore Volcanic Group (Fig. \ref{fig:Geologic_map}). Ref. \citealp{Kulakov2013a} obtained an average VDM of 59 $\pm$ 11 ZAm$^2$ using the \textit{ca.} 1085 Ma Lake Shore Traps volcanics and suggest that the present-day-like geomagnetic field strength likely existed through the late stage of Midcontinent Rift magmatism \cite{Vervoort2007a, Miller2013a}. However, non-ideal paleointensity behaviors at specimen level and overall low success rates have challenged the interpretation of many previous paleointensity results obtained by ref. \citealp{Pesonen1983a, Kulakov2013a}. One typical failure of paleointensity experiment is the double-slope behavior, as two distinct paleointensity estimates may be calculated depending on the interpreter's choice of slope: a higher paleointensity estimate from the low-temperature portion and a lower paleointensity estimate from the high-temperature portion. Ref \citealp{Pesonen1983a} used the low-temperature slope as the best representation of the past magnetic field strength (likely overestimating the field strength) whereas ref. \citealp{Kulakov2013a} used the high-temperature slope (likely underestimating the field strength). Results with such non-ideal behavior were rejected by ref. \citealp{Sprain2018a} who applied stricter paleointensity selection criteria on new paleointensity results and accepted only 23 out of 133 specimens (5 out of 21 lava flows) from the Osler Volcanics and the Mamainse Point volcanics. A calculated mean VDM value of 56 $\pm$ 21 ZAm$^2$ is similar to the average value of the past 300 million years \cite{Sprain2018a}.

Although it is typical of paleointensity experiments to have low success rates, bulk plagioclase-rich rocks \cite{Selkin2000a} and single plagioclase crystals \cite{Tarduno2005a} are found to be able to produce high quality paleointensity results. It has been suggested that plagioclase crystals can protect magnetic minerals that exsolved within them against post-formation weathering and thermochemical alteration during paleointensity experiments \cite{Tarduno2005a}. In this study, we present new paleointensity results from IZZ modified absolute paleointensity experiments \cite{Yu2004a} on a suite of bulk samples from \textit{ca.} 1092 Ma anorthosite xenoliths and their Beaver River diabase host rocks of the Beaver Bay Complex within the Midcontinent Rift. Paleointensity experiments on the anorthosite xenoliths have a very high success rate. Their consistent specimen- and site-level paleointensity results together with rock magnetic data reveal that the anorthosite xenoliths have low anisotropy of thermal remanent magnetization (TRM) and can acquire TRM linearly. Magnetic imaging shows the anorthosite specimens have dominant magnetic carriers interstitial to and within plagioclase crystals and they do not display strong preferred orientations. Our single-slope, high quality paleointensity data add to the reliable paleointensity records during the $\sim$25 Myr of Midcontinent Rift magmatic activity. Our new data from the Midcontinent Rift intrusive rocks support previous data from the volcanics that there was a strong late Mesoproterozoic geodynamo.

@@

The Mesoproterozoic North American Midcontinent Rift is an intracontinental rift that did not result in the break-up of Laurentia. Although the protracted magmatic activity in the rift lasted form \textit{ca.} 1109 Ma to \textit{ca.} 1084 Ma \cite{Swanson-Hysell2019a}, its activity during the main magmatic stage ($\sim$1098–1090 Ma; refs. \citealp{Vervoort2007a, Miller2013a}) was characterized by rapid and voluminous emplacement of intrusive rocks and extrusive volcanics. The $<$500 Kyr emplacement of the \textit{ca.} 1096 Ma Duluth Complex and the associated $\sim$8 km thick extrusive volcanics is one manifestation \cite{Swanson-Hysell2021a}. The emplacement of the \textit{ca.} 1092 Ma Beaver Bay Complex in northern Minnesota punctuates another such period of magmatism during the Midcontinent Rift main stage activity. The hypabyssal intrusion of the 1091.7 $\pm$ 0.2 Ma Beaver River diabase dikes and sills of the Beaver Bay Complex transported numerous anorthosite xenoliths that have short-axis diameters up to 180 meters via wide conduits \cite{Boerboom2004a, Boerboom2006b}. Geochronology, geochemical, and paleomagnetic data support the hypothesis that the diabase extruded to the surface and fed the coeval voluminous Greenstone Flow of the Portage Lake volcanics \cite{Doyle2016a, Zhang2021b}.

Thermal modeling results and paleomagnetic directional data show the anorthosite xenoliths acquired thermal remanent magnetizations during cooling within the Beaver River diabase \cite{Zhang2021b}. Step-wise thermal demagnetizations show the anorthosite xenoliths have minimal secondary remanence and single-component characteristic remanence magnetizations that often unblock sharply within temperature ranges between 500\textdegree C and 580\textdegree C. Geochronology data and rare earth element diffusion modeling further suggest that the anorthosite xenoliths are cumulate enclaves that formed at the time of Beaver Bay Complex magmatism. Petrological textures and geochemical attributes imply that the Beaver River anorthosite xenoliths formed in a lower crustal magmatic conditionm \cite{Zhang2021b, Miller1997a} distinct from that of layered mafic intrusions, such as the Duluth Complex which is $\sim$4 Myr older and generated massive anorthositic cumulates that led to the formation of the Anorthositic Series \cite{Swanson-Hysell2021a, Miller1990a}. The dominantly monomineralic anorthosite xenoliths of the Beaver River diabase have an average anorthite content of $\sim$70\% \cite{Morrison1983a, Doyle2016a}, higher than the average of $\sim$60\% of the Anorthositic Series rocks of the Duluth Complex \cite{Miller1990a}. The plagioclase in the Duluth Complex Anorthositic Series typically are euhedral with an interlocking texture showing igneous foliations (Fig. \ref{fig:Petro_QDM}A), but the more calcium-rich plagioclase of the Beaver River anorthosites often develop granoblastic texture characterized by equigranular crystals with poor alignments (Fig. \ref{fig:Petro_QDM}B). Abundant magnetite and ilmenite needles that are typically 10 $\mu$m long can be found to have preferred orientations within Duluth Complex plagioclase. Interstitial titanomagnetite and ilmenite symplectic intergrowths also coexist with pyroxene and relict olivine as a product of olivine oxidation in the Duluth Complex plagioclase \cite{Miller1990a}. In contrast, titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent but relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{fig:Petro_QDM}). This reflects that the Beaver River diabase magma had a different oxidation history than that of the Duluth Complex. The plagioclase and oxide mineral textures of the Beaver River anorthosites also distinguish them from plagioclase cumulates of other layered mafic complex where magnetic mineral fabrics associated with igneous foliation often occur \cite{Scofield1986a, Selkin2000a, Feinberg2006a}.

% Such a high concentration of Fe oxides carrying magnetic remanence in the slowly cooled Duluth Complex anorthositic rocks reflect that iron in the plagioclase mush in the magma chamber acted as a means of buffering the melt with respect to oxygen, forming crystallites along preexisting plagioclase substrate or as iron oxides incorporated into the plagioclase that subsequently exsolved at subsolidus temperatures \cite{Scofield1986a}. In contrast, titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent but relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{fig:Petro_QDM}). 
% the Beaver River diabase rapidly ascended and entrained the anorthosite xenoliths directly into shallow chambers and likely did not reach  \cite{Flower1980a, Miller1997a, Zhang2021b}. Particularly, \cite{Miller1997a} observed a high state of structural disorder within plagioclase of the Beaver River anorthosite using X-ray diffraction analyses. This is consistent with the interpretation that these anorthosite were rapidly transported by upward currents into hypabyssal magma chambers without reaching structural equilibrium. Within the Beaver River anorthosite xenoliths, the titanomagnetite needles and the symplectic magnetite-ilmenite intergrowths are absent (Fig. \ref{Petro_QDM}). Instead, relatively sparse titano-magnetite within or interstitial to the plagioclase are typical of Beaver River anorthosites (Fig. \ref{Petro_QDM}).

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Geologic_map.pdf}
\caption{\footnotesize{Simplified geologic map of Lake Superior region showing the distribution of rift and post-rift-related rocks including extrusive, intrusive, and late/post-rift sediments around Lake Superior. Red diamonds mark sites with paleointensity results that passed the selection criteria from this study. Paleomagnetic sites from ref. \citealp{Pesonen1983a} are categorized by lithology. All sites from refs. \citealp{Pesonen1983a, Kulakov2013a, Sprain2018a} are color-coded by their ages.}}
\label{fig:Geologic_map}
\end{figure*}

@@

\begin{figure*}
\centering
\noindent\includegraphics[width=11.4 cm]{Petro_QDM.png}
\caption{\footnotesize{Thin section petrographic and magnetic images of anorthosites from Duluth Complex Anorthositic Series (left column) and Beaver River anorthosite xenolith from which paleomagnetic site AX16 and geochronology sample MS99033 were collected (right column; ref. \citealp{Zhang2021b}). Petrographic image (A) shows Duluth Complex anorthosite have euhedral, interlocking plagioclase crystals showing igneous foliation. (B) Beaver River anorthosite have plagioclase with a granoblastic texture. (C) Duluth Complex plagioclase can contain abundant Fe-Ti oxide needles that have preferred orientations that are often parallel with the (010) planes of plagioclase. In contrast, plagioclase of sample MS99033 do not enclose large oxide needles. Magnetic field maps acquired by the quantum diamond microscope (QDM) at UC Berkeley Paleomagnetism Lab show that the oxide needles within Duluth Complex plagioclase are strongly magnetic as evidenced through the complex overlapping magnetic fields exerted above the oxides in (E). In comparison, relatively weak magnetic sources within and interstitial to plagioclase crystals exist in the Beaver River anorthosite. The dipole-like features in (F) do not show strong preferred orientations. Magnetic field maps of (E) and (F) share the same spatial and magnetic scale. The unit of the axes are in $\mu$m.}}
\label{fig:Petro_QDM}
\end{figure*}

@@

40 from a total of 86 anorthosite specimens and 0 out of a total of 69 diabase specimens passed our paleointensity selection criteria, resulting in 7 anorthosite sites and no diabase sites with site-level paleointensity results that pass paleointensity selections. Example NRM/TRM plots are shown in Fig. \ref{fig:IZZI_examples}. Summary specimen absolute paleointensity estimates and site-level weighted mean paleointensity values are shown in Table S1 and are plotted in Fig. \ref{fig:PINT_cooling_corrected}. All measurement-level paleointensity experiment data can be accessed through the MagIC database (\url{}; UPDATE WHEN DOI IS GENERATED).

\begin{figure*}
\noindent\includegraphics[width=17.8cm]{IZZI_examples.pdf}
\centering
\caption{\footnotesize{Example results of paleointensity experiments displayed on NRM/TRM plots and zero-field heating results are shown on orthogonal plots (insets) for anorthosite and diabase specimens. Red (blue) circles indicate zero-field/in-field (in-field/zero-field) steps `ZI’ (`IZ’). Triangles mark partial thermal remanent (pTRM) checks. Blue and red squares in the Zijderveld plots are X–Y and X–Z projections, respectively, of the NRMs in specimen coordinates. Plots on the top row show successful specimen paleointensity results with straight, single-slope behaviors that pass our selection criteria. The green lines represent our fitting for the dominant single-slope component that passes our acceptance criteria with estimates of the ancient field (B$_{anc}$) being used to determine site means. The plots for anorthosite specimens AX1-1a and diabase BD5-1b on the bottom row show typical non-ideal sagging and zigzagging behaviour that fails our acceptance criteria. Plot for diabase specimen BD2-5a shows typical paleointensity behavior of this site - specimens typically have double-slope behaviors although the lower temperature slopes can pass the selection criteria, resulting in the high estimated B$_{anc}$ values shown in Fig. \ref{fig:PINT_cooling_corrected}. Thus, the estimates of B$_{anc}$ for this specimen plot is illustrative, but were not used to calculate a site mean paleointensity value in Fig. \ref{fig:PINT_cooling_corrected}.}}
\label{fig:IZZI_examples}
\end{figure*}

The Beaver River anorthosite specimens yielded a high pass rate of $\sim$46.5\% through our selection criteria. In contrast, the diabase specimens are all rejected. Paleointensity experiment results from both lithologies show three types of dominant specimen behaviors on NRM/TRM plots (Fig. \ref{fig:IZZI_examples}): dominantly single-slope, double-slope, and zigzagging or sagging. 
% Note that almost all specimens have minimal secondary remanence component in thermal demagnetization results as shown in example orthogonal plots of all zero-field steps (Fig. \ref{fig:IZZI_examples}). This reflects that the rocks have experienced mild alterations after their formation. 
Typical specimen paleointensity behavior of the anorthosites display straight, single-slope NRM/TRM plots and the accepted fractions of temperature steps span over the origin-trending, primary remanence components (Fig. \ref{fig:IZZI_examples}). We accept all specimen- and site-level absolute paleointensity estimates from the anorthosite xenoliths that pass the selection. All diabase specimens failed the selection because of double-slope behavior (fail FRAC selection; see methods), poor pTRM checks, and zigzagging (fail SCAT, DANG selection; see methods) plots (Fig. \ref{fig:IZZI_examples}). As is shown in Fig. \ref{fig:IZZI_examples}, specimen BD2-5b has a high-temperature slope with zigzagging behavior, poor pTRM checks, and low NRM/TRM ratio, in addition to a low-temperature slope with a steep NRM/TRM ratio. Although the BD2 specimens' lower temperature fraction of the plots are straight and the pTRM checks are consistent, lower temperature slopes from BD2 give consistently higher paleointensity estimates than those from site AX6, AX8, and AX10, which are anorthosite xenoliths included within BD2 (Fig. \ref{fig:Geologic_map}; ref. \citealp{Zhang2021a}. This higher estimate from the diabase lithology could be attributed to the systematic bias of using the lower temperature slope of the NRM/TRM plots demonstrated in Fig. \ref{fig:IZZI_examples} and specimens from site BD2 all fail the criteria of having a maximum magnetization gap between adjacent steps of 25\% of the total NRM. Eventually, we only accept the single-slope anorthosite specimens results for further analyses.
